---
title: "Piping"
output: github_document
---

```{r, results='hide', echo = FALSE, warning=FALSE, message=FALSE}
library(tidyverse)
library(here)
```

```{r, results='hide', echo = FALSE, warning=FALSE, message=FALSE}
spotify <- read_csv(here("data", "spotify.csv"))
```

[<<< Previous](02_isolating-data.md) | [Next >>>](04_summarizing-data.md)

-----

## %>%

### Steps

Notice how each dplyr function takes a data frame as input and returns a data frame as output. This makes the functions easy to use in a step by step fashion. For example, you could:

1. Filter `spotify` to just songs from before the COVID pandemic, then
2. Select the `loudness` and `energy` columns from the result

```{r echo = TRUE}
pre_covid <- filter(spotify, covid_period == "pre_covid")
pre_covid <- select(pre_covid, loudness, energy)
pre_covid
```

### Redundancy
 Do you notice how we re-create `pre_covid` at each step so we will have something to pass to the next step? This is an inefficient way to write R code.

You could avoid creating `pre_covid` by nesting your functions inside of each other, but this creates code that is hard to read:

```{r echo = TRUE, eval = FALSE}
select(filter(spotify, covid_period == "pre_covid"), loudness, energy)
```

The dplyr package provides a third way to write sequences of functions: the pipe.

### %>%

![Source: dplyr cheatsheet][5]

[5]: images/pipes.png

The pipe operator `%>%` performs an extremely simple task: it passes the result on its left into the first argument of the function on its right. Or put another way, `x %>% f(y)` is the same as `f(x, y)`. This piece of code punctuation makes it easy to write and read series of functions that are applied in a step by step way. For example, we can use the pipe to rewrite our code above:

```{r echo = TRUE}
spotify %>% 
  filter(covid_period == "pre_covid") %>% 
  select(loudness, energy)
```

As you read the code, pronounce `%>%` as "then". You'll notice that `dplyr` makes it easy to read pipes. Each function name is a verb, so our code resembles the statement, "Take spotify, _then_ filter it by mode, _then_ select the loudness and energy."

`dplyr` also makes it easy to write pipes. Each `dplyr` function returns a data frame that can be piped into another `dplyr` function, which will accept the data frame as its first argument. In fact, `dplyr` functions are written with pipes in mind: each function does one simple task. `dplyr` expects you to use pipes to combine these simple tasks to produce sophisticated results.

**Exercise 1**

We'll use pipes for the remainder of the tutorial. Let's practice a little by writing a new pipe. The pipe should:

1. Filter `spotify` to just the songs that are above 0.50 `danceability`
2. Select the `tempo` and `energy` columns

-----

## Using Color

Color and shapes do more than make your plot look nicer. They add a third dimension to your plot that can highlight groups, emphasize higher or lower continuous values, and more. Used wisely, your data visualization can become much more impactful. Used unwisely, you may end up with an ugly plot or, even worse, mislead the viewer. We'll learn the mechanics of how to add color in `ggplot` here, but make sure to check out [this chapter](https://clauswilke.com/dataviz/color-basics.html) in Claus Wilke's Fundamentals of Data Visualization for a more in-depth discussion of using color in your data viz. 

Remember this histogram of valence for pre-COVID songs from the last section?  
```{r, echo=FALSE}
# first filter for pre-covid songs
pre_covid <- filter(spotify, covid_period == "pre_covid")

# now, visualize a histogram of the valence for these songs
ggplot(data = pre_covid, aes(x = valence)) +
  geom_histogram(binwidth = 0.1)
```

To answer the question "Are songs from before the pandemic more positive than those after the pandemic?" I think it's more useful to visualize the two distributions side-by-side than make individual plots for both. One way to do this is to use color.  

To change the color of a plot, we supply either the `color` argument or the `fill` argument. The `color` argument refers to the color of points, lines, and the edges of shapes, while the `fill` argument refers to the color of the inside of shapes (e.g. the inside of a polygon).  
When you use the `color` or `fill` arguments inside of the `aes()` function, you supply the variable to map the colors to. For instance, to split a histogram into two colors, one for the pre-COVID time period, and one for the post-COVID time period, we specify the `covid_period` variable for the `fill` argument.  

```{r}
spotify %>%
  ggplot(aes(x = valence, fill = covid_period)) +
  geom_histogram(binwidth = 0.1)
```

However, this plot isn't super helpful if we're interested in differences in the shape of the distributions. The histogram uses absolute counts, while we want to see a relative comparison. Fortunately, the `geom_density` argument allows us to visualize the relative shapes of two distributions. To visualize the differences between the pre-COVID and post-COVID time periods, we specify the `color` variable.  

```{r}
spotify %>%
  ggplot(aes(x = valence, color = covid_period)) +
  geom_density()
```

Man, it looks like the American public consistently likes to listen to moderately positive music! There is a bit of a jump at the lower range of valence values for the post_COVID time frame, but I'm hesitant to over-interpret that result.  

Let's spruce the plot up a bit by adding a plot title, capitalizing the axis title names, and customizing the legend title. To do this, we use the `labs()` function! The arguments are mostly self-explanatory, except for the legend. To change the title of a legend, you specify the aesthetic the legend is referring to, in this case the `color` aesthetic.  

```{r}
spotify %>%
  ggplot(aes(x = valence, color = covid_period)) +
  geom_density() +
  labs(
    title = "No trend in valence before and after the COVID pandemic",
    x = "Valence",
    y = "Density",
    color = "COVID period"
  )
```

For fun, let's update the color to something more visually appealing. `ggplot2` is notorious for its ugly default color scheme. In our case, let's use the *viridis* color scheme. It's functional and visually appealing, and fortunately comes with its own function! We will use the `scale_color_viridis_d()` function to specify that we want to *scale* the *color* aesthetic using the *viridis* color palette, and the variable we are mapping to is *d*iscrete.  

```{r}
spotify %>%
  ggplot(aes(x = valence, color = covid_period)) +
  geom_density() +
  labs(
    title = "No trend in valence before and after the COVID pandemic",
    x = "Valence",
    y = "Density",
    color = "COVID period"
  ) +
  scale_color_viridis_d()
```


The `scale_*` family of functions maps colors, shapes, transparency, and more to your aesthetics (arguments you provide in the `aes()` function). We used `scale_color_viridis_d()` to map the viridis color palette to the color aesthetic. If we wanted to map the viridis color palette to the `fill` aesthetic, we would use the `scale_fill_viridis_d()` for a discrete variable, or `scale_fill_viridis_c()` for a continuous variable. There are a TON of `scale_*` functions, and fortunately a kind soul recently provided a comprehensive and clear [guide to all of them](https://ggplot2tor.com/scales/). They even made it easy to choose which function you need, given what you want to change!  

**Exercise 2**  
Make a density plot of the `loudness` variable and color it according to the `covid_period` variable. Change the x-axis and y-axis labels so their first letters are upper case and change the legend to read "COVID period". Finally, use one of the `scale_*_viridis_*()` functions to use the viridis color palette instead of `ggplot2`'s default. Add a title describing the plot if you want! What trend (if any) do you see?  
*Extension:* if you're feeling ambitious, try out one of the other `scale_*` functions to color your plot.  

-----

### Go further
There are some fun color palettes out there! Check out these palettes to up your plotting game:  

* [ggpomological](https://www.garrickadenbuie.com/project/ggpomological/) for fruit themed colors  
* [wesanderson](https://github.com/karthik/wesanderson) for colors derived from Wes Anderson movies  
* [ggsci](https://cran.r-project.org/web/packages/ggsci/vignettes/ggsci.html) for color palettes inspired by colors used in scientific journals, data visualization libraries, science fiction movies, and TV shows  

In addition, it's important to make your color choices accessible to those with vision impairments. To see how your color palette is viewed under a range of color blindness syndromes, try out your plots with the [colorblindr](https://github.com/clauswilke/colorblindr) package!  


-----

## Answers

**Exercise 1**  

```{r}
spotify %>% 
  filter(danceability > 0.50) %>% 
  select(tempo, energy) 
```

**Exercise 2**
```{r}
spotify %>%
  ggplot(aes(x = loudness, color = covid_period)) +
  geom_density() +
  labs(
    title = "Slight decrease in loudness after the pandemic began",
    x = "Loudness",
    y = "Density",
    color = "COVID period"
  ) +
  scale_color_viridis_d()
```

Now I'll try it with one of the `scale_color_brewer()` palettes!
```{r}
spotify %>%
  ggplot(aes(x = loudness, color = covid_period)) +
  geom_density() +
  labs(
    title = "Slight decrease in loudness after the pandemic began",
    x = "Loudness",
    y = "Density",
    color = "COVID period"
  ) +
  scale_color_brewer(palette = "Dark2")
```



-----

[<<< Previous](02_isolating-data.md) | [Next >>>](04_summarizing-data.md)
