---
title: "Summarizing data"
output: github_document
---

```{r, results='hide', echo = FALSE, warning=FALSE, message=FALSE}
library(tidyverse)
```

```{r, results='hide', echo = FALSE, warning=FALSE, message=FALSE}
spotify <- read_csv("data/spotify.csv")
```

[<<< Previous](03_piping.md)

-----

## summarise()

![Source: dplyr cheatsheet][6]

[6]: images/summarise.png

Every data frame that you meet implies more information than it displays. For example, `spotify` does not display the average energy of a genre, but `spotify` certainly implies what that number is. To discover the number, you only need to do a calculation:

```{r echo = TRUE}
spotify %>% 
  filter(genre == "Rock") %>% 
  summarise(avg_energy = mean(energy))
```

`summarise()` takes a data frame and uses it to calculate a new data frame of summary statistics. 

### Syntax

To use `summarise()`, pass it a data frame and then one or more named arguments. Each named argument should be set to an R expression that generates a single value. Summarise will turn each named argument into a column in the new data frame. The name of each argument will become the column name, and the value returned by the argument will become the column contents.

### Example

I used `summarise()` above to calculate the average energy of songs in the "Rock" genre, but let's expand that code to also calculate 

* `max` - the rock song with the most energy 
* `sum` - the total energy of all rock songs in the data set

```{r echo = TRUE}
spotify %>%
  filter(genre == "Rock") %>%
  summarise(
    avg_energy = mean(energy),
    max_energy = max(energy),
    total_energy = sum(energy)
  )
```

Wow. Look at that efficient use of pipes (`%>%`) AND `summarise()`!

**Exercise 1**

Compute these three statistics:

* the average loudness of blues songs (`mean()`)
* the maximum danceability of any blues song (`max()`)
* the minimum energy of any blues song (`min()`)

Remember, you will have to use `filter()` to filter for blues songs!

-----

### summarise by groups?

How can we apply `summarise()` to find the average energy for each genre in `spotify`? You've seen how to calculate the average energy of a genre, which gives us the answer for a single genre of interest:

```{r}
spotify %>% 
  filter(genre == "Blues") %>% 
  summarise(avg_energy = mean(energy))
```

However, we had to isolate the genre from the rest of the data to calculate this number. You could imagine writing a program that goes through each genre one at a time and: 

1. filters out the rows with just that genre
2. applies summarise to the rows

Eventually, the program could combine all of the results back into a single data set. However, you don't need to write such a program; this is the job of dplyr's `group_by()` function.

## group_by()

![Source: dplyr cheatsheet][7]

[7]: images/group_by_1.png

`group_by()` takes a data frame and then the names of one or more columns in the data frame. It returns a copy of the data frame that has been "grouped" into sets of rows that share identical combinations of values in the specified columns. 

### group_by() in action

For example, the result below is grouped into rows that have the same genre.

```{r echo = TRUE}
spotify %>%
  group_by(genre)
```

### Using group_by()

![Source: dplyr cheatsheet][8]

[8]: images/group_by_2.png

By itself, `group_by()` doesn't do much. It assigns grouping criteria that is stored as metadata alongside the original data set. If your dataset is a tibble, as above, R will tell you that the data is grouped at the top of the tibble display. In all other aspects, the data looks the same.

However, when you apply a dplyr function like `summarise()` to grouped data, dplyr will execute the function in a groupwise manner. Instead of computing a single summary for the entire data set, dplyr will compute individual summaries for each group and return them as a single data frame. The data frame will contain the summary columns as well as the columns in the grouping criteria, which makes the result decipherable:

```{r}
spotify %>%
  group_by(genre) %>% 
  summarise(min_loud = min(loudness))
```

To understand exactly what `group_by()` is doing, remove the line `group_by(genre) %>%` from the code above and rerun it. How do the results change?

**Exercise 2**
How would you go about finding the 

### Go further
The `mutate()` function is another highly useful tool for extracting unseen insights from your dataframe. While `select()` allows you to choose columns and `group_by()` allows you to summarise rows, `mutate()` enables you to perform calculations and other manipulations on all data within a column. For instance, you're able perform arithmetic on numeric columns or create a new column that is a combination of two existing columns. For a light introduction, [check this out](https://dplyr.tidyverse.org/reference/mutate.html). 

## Put it all together

**
Woo! You now have the tools necessary to answer our original question: Which of your favorite music genres will pep you up the most? To answer this question, you need to:

1. select the columns you need (`genre` and `energy`)
2. filter for your **4** genres of interest
3. group the genres
4. calculate the mean energy for each genre

First, I'll show you what genres you have to work with. Although it's not a focus of this workshop, the `unique()` function is quite useful when you want to know the unique values of a variable.
```{r}
unique(spotify$genre)
```


Now time to try it out! 


-----

### Recap
Congratulations! You can use dplyr's grammar of data manipulation to access any data associated with a table---even if that data is not currently displayed by the table.

In other words, you now know how to look at data in R, as well as how to access specific values and calculate summary statistics.

## Answers

**Exercise 1**

```{r}
spotify %>%
  filter(genre == "Blues") %>%
  summarise(
    avg_loud = mean(loudness),
    max_dance = max(danceability),
    min_energy = min(energy)
  )
```


**Exercise 2**

Using the `|` (or) Boolean operator
```{r}
spotify %>% 
  select(genre, energy) %>% 
  filter(genre == "Ska" | genre == "Rock" | genre == "Rap" | genre == "Indie") %>% 
  group_by(genre) %>% 
  summarize(avg_energy = mean(energy)) 
```


Using the `%in%` Boolean operator
```{r}
spotify %>% 
  select(genre, energy) %>% 
  filter(genre %in% c("Ska", "Rock", "Rap", "Indie")) %>% 
  group_by(genre) %>% 
  summarize(avg_energy = mean(energy)) 
```


-----

[<<< Previous](03_piping.md)

